```{r, echo=FALSE, results='hide'}
source('/Users/sarah/Documents/GitHub/experimental-rads/expRADs_md_script.r')
```

**Title:** Assessing the impact of perturbations on taxonomic biodiversity using a cross-community approach
--------------------------------------------------------
**Authors:** Sarah R. Supp and S. K. Morgan Ernest

**Affiliations:** Utah State University, Department of Biology and the Ecology Center
5305 Old Main Hill, Logan, UT 84322-5305

The authors declare no conflict of interest

**Corresponding author information:** Address: Sarah R. Supp, 5305 Old Main Hill, Logan, UT, 84322-5305; Email: sarah.supp@usu.edu



**Abstract:**
Ecologists use a wide variety of biodiversity metrics to study the abundance and distribution of species. While it is increasingly clear that researchers will need to forecast changes in biodiversity, ecology currently lacks a framework for understanding the natural background variability in biodiversity or how biodiversity patterns will respond under environmental change scenarios. The most commonly used biodiversity metrics include species composition, species abundance, species richness, and the species abundance distribution (SAD) that incorporates the metrics of species richness, abundance, and evenness. Together, these metrics are used to describe biodiversity across ecosystems and taxonomic groups, but their sensitivity to environmental changes is unknown. Using global-span data from small-scale terrestrial animal communities, we show that the SAD and species richness are generally resilient under a suite of artificial and natural manipulations.  In contrast, species composition and abundance responded readily to manipulation. Our results suggest that species richness and the SAD are poor indicators of change, perhaps because many systems display strong compensatory dynamics. An important next step in biodiversity research is to place experimental results in a realistic context with respect to levels of background variability in species richness and to further compare the response of different metrics to manipulations. 

**Keywords:** biodiversity, species abundance distribution, disturbance, resilience, community structure



**Introduction**
In an era of rapid natural and anthropogenic change, it is increasingly clear that ecology needs to do more than quantify biodiversity for a snapshot in time. Ecology must also be able to forecast changes in biodiversity for systems in flux (Thomas et al. 2004, Araujo & Rahbek 2006, Dawson et al. 2011). Predicting biodiversity in disturbed systems has several challenges, including the need to understand the network of processes driving biodiversity and the ability to predict changes of multiple components of biodiversity under global environmental change scenarios (Fisher et al. 2010, White et al. 2010). Disturbance itself is a fundamental process that can enhance or reduce biodiversity (Hooper et al. 2005, Dornelas et al. 2011). Since disturbance frequency and intensity (e.g. fire, flood, fragmentation) may be altered under global change scenarios, a general framework for understanding biodiversity response to perturbation is critical. 

Biodiversity is a concept that encompasses many different dimensions, including taxonomic, genetic, phylogenetic, and functional (Magurran & McGill 2011). Even within one of these dimensions, diversity is often multifaceted and no single metric quantifies the different aspects researchers are interested in. For example, for taxonomic diversity we are often interested not only in the raw number of species present but also in patterns of dominance and rarity (Magurran 2004, Magurran & McGill 2011). The most commonly used metrics of taxonomic diversity include species composition, the number of individuals, species richness, and species evenness (Dornelas et al. 2011, Maurer & McGill 2011). The species abundance distribution (SAD) is another commonly used metric that incorporates richness, abundance, and evenness to describe overall community structure (e.g., Magurran & Henderson 2003, McGill et al. 2007, Ulrich et al. 2010). Because each metric differs in how it describes diversity, they may also differ in their sensitivity to disturbance and ability to detect community responses (Dornelas et al. 2011). For example, the SAD may pick up important changes in the dominance structure within a community, but will not necessarily be sensitive to change in the density of individuals or species identity (refs).  Because taxonomic biodiversity is multi-faceted, a multi-metric approach to studying biodiversity responses to perturbation will be critical for gaining a holistic understanding of biodiversity dynamics (refs). 

One of the current challenges for predicting biodiversity responses to perturbations is a lack of a general understanding of how biodiversity actually responds to disturbances. A suite of studies has attempted to understand the consequences of decreasing taxonomic diversity on community and ecosystem structure, primarily by focusing on reductions in species richness (e.g., e.g. Naeem et al. 1995, Hector et al. 1999, Reich et al. 2012). However, observational studies suggest that, temporally, species richness appears to exhibit low background variability, at least in systems not obviously perturbed (Brown et al. 2001, Goheen et al.2005, Elmendorf & Harrison 2011). Even in disturbed systems, species richness has often been observed to be relatively insensitive to many perturbations (Chapin III et al. 2000, Hillebrand et al. 2008). These case studies raise the possibility that reduced species richness may not be a general response of communities to disturbance, but that perhaps other changes in biodiversity are more likely. This makes it unclear whether the magnitude of diversity reduction seen in experiments reflects diversity reductions that occur in nature in response to natural and anthropogenic perturbations.  
  
Using a compilation of experimentally manipulated community-level data of terrestrial animal taxa from the published literature (Table S1), we evaluated the response of four commonly used biodiversity metrics to environmental change: species composition, total abundance (N), species richness (S), and the SAD. This approach allows us to specifically address 1) whether community perturbations cause shifts in a suite of biodiversity metrics and 2) if community-level biodiversity-metrics exhibit different sensitivities to perturbations.  

**Methods**
*Database compilation*
We conducted a literature search using Google Scholar October 2011 – February 2012. Peer-reviewed articles that included data tables that reported species-level abundance for a control community and at least one manipulated community were recorded. Published data was often summed or averaged over replicates, rather than reporting abundance separately for each replicate. The data were recorded from a wide variety of sites including manipulated, artificial experiments (i.e., caged exclosures, habitat modules, nutrient addition) and human-mediated “natural” experiments (i.e., controlled burn, silvicultural treatment, grazed plots). Sites represent all continents except Antarctica, and widely varying terrestrial animal taxa (e.g., zooplankton, arachnid, mammal, bird). The data were recorded in linked tables describing the reference, site, experiment, and community details (Appendix A).    

*Data selection*
For analysis, we eliminated communities where > 10% of individuals were not identified to the species level or where the area sampled for paired control – experiment communities was unequal. We used data where raw abundance was reported as a summed total for each species or where mean abundance was reported across the replicates, excluding percent cover, biomass, and presence-only data. In rare cases where mean abundance was reported using less than symbols (e.g. < 0.01) we assumed the value was at the top of that bin. SADs are difficult to characterize when the number of species or total abundance is very low, so we included only communities with S ≥ 5 and N ≥ 30. We compared pairs of communities at sites that were sampled at the same spatial scales and at similar temporal scales (e.g., we did not compare data from different seasons or across a time-series) to avoid complications due to differences in sampling intensity or timing (Gotelli & Colwell 2001, Magurran 2004). Data meeting the criteria was comprised of `r length(unique(cID))` control and `r length(unique(eID))` experiments, representing `r length(Je)` paired control-experiment comparisons from `r length(unique(refID))` published papers. Species richness ranged `r min(CS,ES)`-`r max(CS,ES)` and total abundance ranged `r min(CN,EN)`-`r max(CN,EN)`.

*Characterizing and comparing paired communities*
For each paired community, we compared the relative rank abundance distributions (RAD), species composition, total abundance, and species richness (for plots of all comparisons, Appendix B). RADs are an alternate visualization to SADs where the relative abundance of each species in the community is ordered from most abundant to least abundant. The RAD uses relative abundances and thus minimizes the impacts on the distribution caused primarily by change in total abundance, and also minimizes the information lost using histogram binning methods, especially in smaller communities (McGill et al. 2007).

We used the Bray-Curtis dissimilarity metric (BC) to quantify differences between controls and experiments for each of our biodiversity metrics (vegan, function vegdist; Oksanen et al. 2010). Bray-Curtis dissimilarity is a semi-metric index that provides a dissimilarity measure ranging from 0 (two communities are the same) to 1 (completely different). It is commonly used to examine similarity of ecological communities. When used to examine species composition or the RAD, it takes into account the presence/absence of a given species or rank across the two communities, but also the relative abundance of each species or rank. We also characterized RADs using Simpson’s evenness (J), which describes how similar ranks are in their abundances and is somewhat independent of S and N (Magurran 2004, McGill 2011; vegan, function diversity; Oksanen et al. 2010). For other biodiversity metrics, S and N, we calculated the BC dissimilarity and percent difference for each control-experiment community pair.

*Statistical analysis*
To compare measures of species composition, total abundance, species richness, Simpson's evenness across ranks, and relative abundance at each rank (a proxy for the shape of the rank abundance distribution), we calculated fit of values to the 1:1 line (R2), which represents no change in values from control to experimental manipulation (Fig. 1). We also calculated root mean squared error (rmse), which is used to obtain the standard deviation of model prediction error, where lower values indicate better performance. Here, we used the control data as our “observed” and the experimental data as our “predicted” data in order to determine the degree of change, or variance between N, S, and evenness at paired sites (package hydroGOF, function rmse; Zambrano-Bigiarini 2011).
Data and all necessary code for replicating the results are available online  and in the Supplement. 

**Results**
All communities experienced compositional differences between compared manipulated and control sites (Fig 1a, Fig 2a; R2=`r comp_r2`, rmse=`r comp_rmse`, BC values ranged `r round(min(BCcomp),3)`-`r round(max(BCcomp),3)`). Many communities experienced a change in total abundance (Fig 1b, Fig 2b; R2=`r n_r2`, rmse=`r n_rmse`; BC values ranged `r round(min(BCN),3)`-`r round(max(BCN),3)`). Most communities experienced relatively small magnitude change in species richness (Fig 1c, Fig 2c; R2=`r s_r2`, rmse=`r s_rmse`, BC values ranged `r round(min(BCS),3)`-`r round(max(BCS),3)`). Change in the abundance at each rank for paired RADs (Fig1e, Fig 2e; R2=`r relabun_r2`, rmse=`r relabun_rmse`, BC values ranged `r round(min(BCrad),3)`-`r round(max(BCrad),3)`) and in evenness (Fig 1d, Fig 2d; R2=`r j_r2`, rmse=`r j_rmse`, BC values ranged `r round(min(BCJ),3)`-`r round(max(BCJ),3)`) were also generally small magnitude. 

####**Discussion**
Across a wide range of taxonomic groups, ecosystems, and experimental treatments, community perturbations caused shifts in a suite of biodiversity metrics, but metrics exhibited different sensitivities to perturbations. The small magnitude of changes observed in species richness and the rank abundance distribution suggests that they are relatively insensitive to disturbance, including perturbations that most ecologists would agree constitute as “major” changes to a system (e.g., wildfire, clear-cuts). In contrast, species composition and total abundance exhibited larger responses to perturbations. Our results suggest that the primary response to perturbations is compositional reorganization and a change in the number of individuals supported within a community, with generally moderate to minor changes in the overall structure of the community (i.e., number of species supported and structure of commonness and rarity).

Because disturbances typically change important environmental conditions that can benefit some species to the detriment of others, the strong response of species composition to disturbance is not surprising. Similarly, the strong response of total abundance is not surprising since many disturbances either directly thin communities (e.g., fire) or alter availability of limiting resources (e.g., nutrient additions).  More intriguing is that the reorganization of total abundance and species composition does not also correspond with shifts in species richness or structure of a community.  Recent work on the species abundance distribution showed that this metric can be insensitive to changes in species composition (Supp et al 2012) and our results suggest that this finding may be general. The relatively constrained response of species richness to perturbations suggests that the case studies showing that taxonomic species richness is relatively robust to natural and experimental perturbations is a general phenomenon. For these metrics to be resilient to shifts in total abundance and species composition there must be mechanisms that both maintain the number of species in a community and its pattern of dominance and rarity. Compensatory dynamics have been implicated in other studies studying stability of species richness and may be an important mechanism mitigating the effects of species loss in compositionally reorganized systems in our study. If sites maintain adequate connectivity with regional species pools that enable compensatory dynamics through immigration or if the species pool is relatively large (Chao & Jost 2012), high magnitude changes in species richness and the SAD may be relatively rare (e.g., Ernest & Brown 2001, Gonzalez & Loreau 2009, Dornelas 2010). Because species richness and the form of the SAD are relatively resilient to disturbance, we suggest that when high magnitude shifts do occur they may indicate fundamental changes that warrant further study and attention. Critical conservation decisions for threatened areas are often made based on continuing function of ecosystem services (e.g., Chan et al. 2006, Nelson et al. 2009), which in turn are often dependent on maintaining specific biological interactions. As such, the details of biotic interactions are often quite important and high magnitude changes in species composition and abundance cannot be ignored. 

Species richness is often used to quantify the influence of disturbance on a community or ecosystem (Costello et al. 2004, Dornelas et al. 2011), but our results and others (e.g., Chapin III et al. 2000, Hillebrand et al. 2008) suggest that it is often surprisingly insensitive to the changes applied to a system. Decades of diversity experiments have manipulated species richness to understand the impacts of species gain or loss on ecosystems (e.g. Naeem et al. 1995, Hector et al. 1999, Reich et al. 2012), but ecologists still generally lack an understanding of the existing background variability of species richness at natural sites, the magnitude of changes that occur in response to disturbance, or what magnitude of change in richness is biologically significant. In our analysis, only one site experienced more than a twofold change in species richness. This raises the important question of what diversity experiments actually reveal about ecological response to disturbance if the magnitude of species richness change studied in these systems is often much larger than what is empirically observed.

A major challenge facing biodiversity research is to bridge the existing gap between basic and applied ecological research. Our results suggest that biodiversity response to disturbance is complex and highly dependent on which component of biodiversity is being studied. More research is needed to evaluate the sensitivity of different biodiversity metrics to environmental change and the background variability in biodiversity in relatively undisturbed systems in order to develop a dynamic framework for better prediction. Future research placing biodiversity experiments in the context of realistic rates of change in ecological time is an important next step forward. 

###**Acknowledgements.** We thank E.P. White, P.B. Adler, D.N. Koons, M.E. Pfrender, B.J. McGill, and the Weecology Group at Utah State University for their statistical advice and discussions that greatly improved the manuscript.  We thank all the authors who collected and published the data. SRS was supported by USU Ecology Center and School of Graduate Studies fellowships.

Figures and Tables:
-------------------

**Figure 1.** Panel of 1:1 plots with R2 for a) species composition relative abundance b) total abundance, c) species richness, d) rank evenness, and d) relative abundance at each rank of the RAD. Control data is on the x-axis and experimental data is on the y-axis. Fit to the 1:1 line (dashed line) suggests no change in the parameter among the paired control-experiment comparison. 

```{r fig.width=3, fig.height=10, echo=FALSE}
par(mfrow=c(5,1), mar=c(1.5,1.5,2,0.5), oma=c(1.5,2,1,1)) 

# Fig 1A - plots for Composition
plot(NA, NA, xlim = c(0,1), ylim = c(0,1), xlab = "", ylab = "", bty = "n") 
abline(0, 1, lty = 2, col = 'gray20', lwd = 1)
points(compc, compe, pch = 19)
#mtext('Species Composition', side = 3, line = -0.25, cex = 0.75)
mtext('relative abundance', side = 2, line = 2, cex = 0.75)
mtext('relative abundance', side = 1, line = 2, cex = 0.75)

# Fig 1B - plots for N
plot(NA, NA, pch = 19, log = 'xy', xlim = c(1,6500), ylim = c(1,6500),
     xlab = '', ylab = '', bty = 'n')
abline(0, 1, lty = 2, lwd = 1, col = 'gray20')
points(CN, EN, pch = 19, xlim = c(30, 6500, ylim = c(30,6500), xlab = "", ylab = "", main = ""))
#mtext('Total Abundance', side = 3, line = -0.25, cex = 0.75)
mtext('total abundance', side = 2, line = 2, cex = 0.75)
mtext('total abundance', side = 1, line = 2, cex = 0.75)

# Fig 1C - plots for S
plot(NA, NA, log = 'xy', pch = 19, xlim = c(1,200), ylim = c(1,200),
     xlab = '', ylab = '', bty = 'n')
abline(0, 1, lty = 2, lwd = 1, col = 'gray20')
points(CS, ES, pch = 19, xlim = c(5,200), ylim = c(5,200), xlab = 'c', ylab = '', main = '')
#mtext('Species Richness', side = 3, line = -0.25, cex = 0.75)
mtext('species richness', side = 2, line = 2, cex = 0.75)
mtext('species richness', side = 1, line = 2, cex = 0.75)

# Fig 1D - plots for evenness
plot(NA, NA, xlim = c(0,1), ylim = c(0,1), pch = 19, xlab = '', ylab = '', bty = "n")
abline(0,1, lty = 2, lwd = 1, col = 'gray20')
points(Jc, Je, pch = 19, xlim = c(0,1), ylim = c(0,1), xlab = '', ylab = '', main = '')
#mtext('Evenness', side = 3, line = -0.25, cex = 0.75)
mtext("Simpson's evenness", side = 2, line = 2, cex = 0.75)
mtext("Simpson's evenness", side = 1, line = 2, cex = 0.75)

#### Fig 1E.  compare relative abundance at each rank in all paired sites
plot(NA, NA, xlim = c(0,1), ylim = c(0,1), xlab = "", ylab = "", bty = "n") 
abline(0, 1, lty = 2, col = 'gray20', lwd = 1)
points(c, e, pch = 19)
#mtext('Relative Rank Abundance', side = 3, line = -0.25, cex = 0.75)
mtext('relative abundance', side = 2, line = 2, cex = 0.75)
mtext('relative abundance', side = 1, line = 2, cex = 0.75)

box("outer", lty = "solid", col = "black")
```

**Figure 2.** Panel showing the change in the biodiversity metrics among the paired control-experiment sites: a) Bray-Curtis dissimilarity in species composition, b) absolute percent change in total abundance, c) absolute percent change in species richness, d) Bray-Curtis dissimilarity of rank evenness, and e) Bray-Curtis dissimilarity of the relative abundance at each rank for compared RADs. 

```{r fig.width=3, fig.height=7.5, echo=FALSE}
par(mfrow=c(5,1), mar=c(3,1.5,2,0.5), oma=c(1.5,2,1,1))

binwidth <- 0.1
#Bray-Curtis dissimilarity: 0 = the same, 1 = completely different"
# Fig 2A. Species composition
hist(BCcomp, xlim = c(0,1), ylim = c(0,50), breaks = seq(0, 1, by = binwidth), main = "",
     col = "gray80", xlab = "", ylab = "")
#    mtext('Composition', side = 3, line = -0.75)
    mtext('composition Bray-Curtis dissimilarity', side = 1, line = 2, cex = 0.75)

# Fig 2B. Abundance
hist(abs(percN), xlim = c(0, 500), ylim = c(0,30), breaks = seq(0, 700, by = 10), main = "",
     col = "gray80", xlab = "", ylab = "")
#    mtext('Total Abundance', side = 3, line = -0.75)
    mtext('total abundance percent difference', side = 1, line = 2, cex = 0.75)

# Fig 2C. Species Richness
hist(abs(percS), xlim = c(0,120), ylim = c(0,40), breaks = seq(0, 120, by = 10), main = "",  
     col = "gray80",  xlab = "", ylab ="")
#    mtext('Species Richness', side = 3, line = -0.75)
    mtext('species richness percent difference', side = 1, line = 2, cex = 0.75)
    mtext('number of paired comparisons', side = 2, line = 2.25, cex = 0.75)

# Fig 2D. Rank Evenness
hist(BCJ, xlim = c(0,1), ylim = c(0,60), breaks = seq(0, 1, by = binwidth), main = "",
     xlab = "", ylab = "", col = "gray80")
#    mtext('Rank Evenness', side = 3, line = -0.75)
    mtext('rank evenness Bray-Curtis dissimilarity', side = 1, line = 2, cex = 0.75)

# Fig 2E. Relative Rank Abundance Dissimilarity
hist(BCrad, xlim = c(0,1), ylim = c(0,60), breaks = seq(0, 1, by = binwidth), main = "", 
     xlab = "", ylab = "", col = "gray80")
#    mtext('Rank Abundance', side = 3, line = -0.75)
    mtext('rank abundance Bray-Curtis dissimilarity', side = 1, line = 2, cex = 0.75)

box("outer", lty="solid", col="black")
```
